---
title: "Diamond"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#DIAMOND:

How many observations and how many variables in the diamonds dataset?
53940 - 10
How many order factors in the variables?
```{r}
library(ggplot2)
data(diamonds)
summary(diamonds)
str(diamonds)
```
what letters represents the best diamond?
```{r}
levels(diamonds$color)
```

#Price Histogram:
Create a histogram of the price of all the diamonds in the diamond data set.

```{r}
library(ggthemes) 
#theme_set(theme_minimal(24)) 
qplot(x = price, data = diamonds)
```
long-tailed distribution.
#Price Histogram Summary:
Describe the shape and center of the price distribution. Include summary statistics like mean and median

```{r}
summary(diamonds$price)
```

# Diamond Counts:
How many diamonds cost less than $500,250 or more than 15000
```{r}
sum(diamonds$price < 500)
sum(diamonds$price <250)
sum(diamonds$price>=15000)

```
# Cheaper Diamonds:
Explore the largest peak in the price histogram you created earlier.
Try limiting the x-axis, altering the bin width, and setting different breaks on the x-axis.
```{r}
qplot(x = price, data = diamonds)
qplot(x = price, data = diamonds, xlim = c(0,2500), binwidth = 1)
qplot(x = price, data = diamonds, xlim = c(0,2500), binwidth = 200)
qplot(x = price, data = diamonds, xlim = c(0,2500), binwidth = 500)
ggsave('priceHistogram.png')
```
There are no diamonds that cost $1500.
For diamonds that cost less than $2,000, the most common price of a diamond is around $700 with the mode being $605 (binwidth = 1). 

# Price by Cut Histograms:
Break out the histogram of diamond prices by cut.
Which cut has the highest priced dimond?

```{r}
qplot(x = price, data = diamonds) + 
  facet_wrap(~cut)

diamonds$cut[which.max(diamonds$price)]
diamonds$cut[which.min(diamonds$price)]
by(diamonds$price, diamonds$cut, summary)
```
we looked at the summary statistics for diamond price by cut. If we look at the output table, the the median and quartiles are reasonably close to each other.This means the distributions should be somewhat similar, but the histograms we created don't show that.The 'Fair' and 'Good' diamonds appear to have different distributions compared to the better cut diamonds. They seem somewhat uniform on the left with long tails on the right. So, we are going to fixed the scaled.

#Scales and Multiple Histograms:
add a parameter to facet_wrap so that the y-axis in the histograms is not fixed. You want the y-axis to be different for each histogram.

```{r}

qplot(x = price, data = diamonds) + 
  facet_wrap(~cut, scales = "free")

```

# Price per Carat by Cut:
Create a histogram of price per carat and facet it by cut.
Adjust the bin width and transform the scale of the x-axis using log10.

```{r}
qplot(x = log10(price/carat+1), data = diamonds, binwidth = 0.3)+
  facet_wrap(~cut, scales = "free")
```
# Price Box Plots:
Investigate the price of diamonds using box plots, numerical summaries, and one of the following categorical variables: cut, clarity, or color.

```{r}
by(diamonds$price, diamonds$cut, summary)
by(diamonds$price, diamonds$clarity, summary)
by(diamonds$price, diamonds$color, summary)
qplot(x = cut, y = price, data = diamonds, geom = 'boxplot')
qplot(x = clarity, y = price, data = diamonds, geom = 'boxplot')
qplot(x = color, y = price, data = diamonds, geom = 'boxplot')
```

# Interquartile Range - IQR
what is the price range for the middle 50% of diamonds with color D? what is the price range for the middle 50% of diamonds with color J? what is the IQR for the diamonds with best color? what is the IQR for the diamonds with worst color?

```{r}
summary(diamonds$price[diamonds$color == "D"])
summary(diamonds$price[diamonds$color == "J"])
IQR(diamonds$price[diamonds$color == "D"])
IQR(diamonds$price[diamonds$color == "J"])
```

#Price per Carat Box Plots by Color:
Investigate the price per carat of diamonds across the different colors of diamonds using boxplots.

```{r}

qplot(x = color, y = price/carat, data = diamonds, geom = 'boxplot')+
    coord_cartesian(ylim = c(0,6000))

```
# Carat Frequency Polygon:

investigate the weight of the diamonds using frequency polygon. what carat size has a count greater than 2000?

```{r}
qplot(x = carat, data = diamonds,binwidth = 0.1,
      geom = "freqpoly")+
  scale_x_continuous(limits = c(0,6), breaks = seq(0,6,0.1))


table(round(diamonds$carat,1))
```

```{r}
require(knitr) # required for knitting from rmd to md
require(markdown) # required for md to html 
knit('Lesson3CaseStudy.rmd', 'Lesson3CaseStudy.md') # creates md file
markdownToHTML('Lesson3CaseStudy.md', 'Lesson3CaseStudy.html') # creates html file
browseURL(paste('file://', file.path(getwd(),'Lesson3CaseStudy.html'), sep='')) 
```

